---
title: "Speed and Accuracy Comparison"
description: >
  Reproducible benchmark comparing default GenzBretz, Miwa, fixed-budget
  original randomized QMC, and the new deterministic QMC backend.
---

```{r setup, include = FALSE}
library("mvtnorm")
```

<style>
table.dqmc-results {
  display: table !important;
  width: 100%;
  table-layout: fixed;
  font-size: 0.78rem;
}
table.dqmc-results th,
table.dqmc-results td {
  padding: 0.22rem 0.35rem;
  vertical-align: top;
  white-space: normal;
  overflow-wrap: anywhere;
}
table.dqmc-results th:nth-child(1),
table.dqmc-results td:nth-child(1) {
  width: 2.5rem;
}
table.dqmc-results th:nth-child(2),
table.dqmc-results td:nth-child(2) {
  width: 11rem;
}
@media (max-width: 768px) {
  table.dqmc-results {
    font-size: 0.68rem;
  }
  table.dqmc-results th,
  table.dqmc-results td {
    padding: 0.18rem 0.22rem;
  }
}
</style>

## Purpose

The default `GenzBretz()` algorithm is randomized quasi-Monte Carlo. Its
randomization is one reason it is fast and useful: the algorithm can combine
low-discrepancy integration points with a stochastic error assessment.

`DeterministicQMC()` removes that randomization. The useful benchmark question
is therefore empirical: how much speed is traded for deterministic
reproducibility, and how close are the resulting probabilities on representative
normal-probability problems?

This article compares eight choices:

- `GenzBretz()`: the current default, with default stopping tolerances.
- `Miwa(steps = 128)`: the established deterministic low-dimensional normal
  algorithm.
- fixed-budget original randomized QMC:
  `GenzBretz(maxpts = 25000, abseps = 0, releps = 0)`.
- `DeterministicQMC(maxpts = 25000, backend = "c")`: the new plain C
  deterministic Halton backend for normal probabilities.
- `DeterministicQMC(maxpts = 25000, backend = "sobol")`: the new plain C
  deterministic Sobol backend for normal probabilities.
- `DeterministicQMC(maxpts = 25000, backend = "hybrid")`: an experimental
  Sobol backend with a deterministic tail-spaced change of variables.
- `DeterministicQMC(maxpts = 25000, backend = "genzsobol")`: an experimental
  Sobol backend with variable reordering and embedded stability stopping.
- `DeterministicQMC(maxpts = 50000, backend = "auto")`: the automatic
  deterministic policy, using cross-checked Genz-Sobol when supported and
  deterministic FORTRAN otherwise.

The fixed-budget original QMC row is separate from the default because
`GenzBretz()` may stop early once its tolerance is met. The fixed-budget row is
the cleaner comparison to `DeterministicQMC()` because both are asked to spend
the same nominal number of points. The Auto row is deliberately given a larger
budget because it splits the work across deterministic cross-checks; this is a
benchmark setting for the requested `abseps = 1e-5`, not a proposed package
default.

## Benchmark design

The examples use Toeplitz correlation matrices, semi-infinite lower bounds, and
finite upper bounds. Dimensions are deliberately modest so the comparison is
fast and so `Miwa()` is still available. Accuracy is reported as relative error
against `Miwa(steps = 512)`, to five significant digits. This is a stable
deterministic yardstick for this small benchmark rather than a mathematical
proof of exactness.

Each reported time is the median elapsed time over three evaluations on the
machine building the site.

```{r benchmark}
make_problem <- function(n) {
    list(corr = toeplitz(0.35^(0:(n - 1))),
         lower = rep(-Inf, n),
         upper = seq(-0.2, 0.6, length.out = n))
}

methods <- list(
    list(name = "Default GenzBretz",
         algorithm = GenzBretz(),
         seed = 11L),
    list(name = "Miwa steps=128",
         algorithm = Miwa(steps = 128),
         seed = NULL),
    list(name = "Original randomized QMC",
         algorithm = GenzBretz(maxpts = 25000, abseps = 0, releps = 0),
         seed = 11L),
    list(name = "DeterministicQMC C",
         algorithm = DeterministicQMC(maxpts = 25000, abseps = 0,
                                      releps = 0, backend = "c"),
         seed = NULL),
    list(name = "DeterministicQMC Sobol",
         algorithm = DeterministicQMC(maxpts = 25000, abseps = 0,
                                      releps = 0, backend = "sobol"),
         seed = NULL),
    list(name = "DeterministicQMC Hybrid",
         algorithm = DeterministicQMC(maxpts = 25000, abseps = 0,
                                      releps = 0, backend = "hybrid"),
         seed = NULL),
    list(name = "DeterministicQMC GenzSobol",
         algorithm = DeterministicQMC(maxpts = 25000, abseps = 1e-5,
                                      releps = 0, backend = "genzsobol"),
         seed = NULL),
    list(name = "DeterministicQMC Auto",
         algorithm = DeterministicQMC(maxpts = 50000, abseps = 1e-5,
                                      releps = 0, backend = "auto"),
         seed = NULL)
)

timed_value <- function(problem, method, reps = 3L) {
    times <- numeric(reps)
    value <- NA_real_
    diagnostic <- NA_real_
    target_requested <- is.list(method$algorithm) &&
        ((length(method$algorithm$abseps) == 1L &&
          !is.na(method$algorithm$abseps) && method$algorithm$abseps > 0) ||
         (length(method$algorithm$releps) == 1L &&
          !is.na(method$algorithm$releps) && method$algorithm$releps > 0))
    target_status <- "not requested"
    for (i in seq_len(reps)) {
        if (!is.null(method$seed))
            set.seed(method$seed)
        invisible(gc(FALSE))
        tm <- system.time({
            ans <- pmvnorm(lower = problem$lower, upper = problem$upper,
                           corr = problem$corr,
                           algorithm = method$algorithm)
        })
        times[i] <- unname(tm["elapsed"])
        if (i == 1L) {
            value <- as.numeric(ans)
            diagnostic <- attr(ans, "error")
            if (target_requested)
                target_status <- if (identical(attr(ans, "msg"),
                                               "Normal Completion"))
                    "met" else "not met"
        }
    }
    list(value = value, diagnostic = diagnostic,
         target_status = target_status, elapsed = median(times))
}

benchmark_dimension <- function(n) {
    problem <- make_problem(n)
    reference <- as.numeric(pmvnorm(lower = problem$lower,
                                    upper = problem$upper,
                                    corr = problem$corr,
                                    algorithm = Miwa(steps = 512)))
    rows <- lapply(methods, function(method) {
        ans <- timed_value(problem, method)
        data.frame(dimension = n,
                   method = method$name,
                   elapsed_sec = sprintf("%.4f", ans$elapsed),
                   value = sprintf("%.8g", ans$value),
                   diagnostic = sprintf("%.3e", ans$diagnostic),
                   target_status = ans$target_status,
                   rel_error = sprintf("%.5g",
                       if (reference == 0) NA_real_ else
                           abs(ans$value - reference) / abs(reference)),
                   stringsAsFactors = FALSE)
    })
    do.call(rbind, rows)
}

results <- do.call(rbind, lapply(c(4L, 8L, 12L), benchmark_dimension))
```

```{r benchmark-results, echo=FALSE, results='asis'}
display_results <- results
names(display_results) <- c("dim", "method", "seconds", "value",
                            "diagnostic", "target", "rel error")
knitr::kable(display_results, format = "html", row.names = FALSE,
             escape = TRUE,
             align = c("r", "l", "r", "r", "r", "l", "r"),
             table.attr = 'class="table-sm dqmc-results"')
```

## Reading the results

The default `GenzBretz()` row answers the practical question: what happens if
the package default is used? The fixed-budget QMC row answers a narrower
question: what does the original randomized QMC path do if it is given the same
nominal point budget as the deterministic C backend?

For low dimensions, `Miwa()` is often competitive and accurate. Its
disadvantage is structural: it is a grid-based deterministic method, it is
limited to normal probabilities, and it is limited to dimension 20.

`DeterministicQMC()` occupies a different point in the design space. It gives a
deterministic QMC-style calculation under `maxpts` control. This is useful when
repeated evaluations must agree without managing seeds. The Halton row is the
simple first C backend. The Sobol row tests the first stronger deterministic
point set in the same transformed integral. The Hybrid row keeps the Sobol
backbone but applies a deterministic tail-spaced warp with a Jacobian
correction. The trade-off is that the C backends do not provide the mature
randomized error assessment or all adaptive refinements of `GenzBretz()`.
The GenzSobol row is allowed to use its deterministic embedded stopping rule,
so it answers a slightly different speed question than the fixed-budget C,
Sobol, and Hybrid rows. The Auto row adds deterministic cross-checks at fixed
sequence offsets and reports the larger of the embedded Genz-Sobol stability
errors and the between-check spread. When a tolerance is requested, those
individual checks use a tighter internal target before the combined diagnostic
is compared with the requested `abseps`/`releps`. It is therefore the closest
row to the unattended best-practice deterministic policy: somewhat more work is
acceptable in exchange for a more conservative reproducible accuracy
diagnostic. The larger Auto budget in this benchmark compensates for those
repeated cross-checks; it should not be read as a change to the current package
defaults.

When `abseps` or `releps` is positive, `DeterministicQMC()` treats it as a
request that the reported deterministic diagnostic meet the requested target
before `maxpts` is exhausted. If the diagnostic is larger than the requested
tolerance, the returned object has `inform = 1` and the usual message
`"Completion with error > abseps"`. In the table, `target_status` records
whether such a target was requested and met. This is a warning about the
diagnostic, not a proof that the true integration error is above or below the
target.

<span id="a-proposed-speed-target"></span>

## Sobol, hybrid, Genz-Sobol

The benchmark now includes three implementations of the proposed deterministic
Genz-Sobol direction. `backend = "sobol"` keeps the sequential Genz
transformation and uses Sobol direction numbers with Gray-code point
generation in plain C. `backend = "hybrid"` applies a gentle deterministic
tail-spaced change of variables to the Sobol coordinates, with the corresponding
Jacobian correction. That makes the nonuniform-grid idea testable without
replacing QMC by a full tensor grid. `backend = "genzsobol"` adds
marginal-width variable reordering and embedded powers-of-two stability checks
using `abseps` and `releps`. `backend = "auto"` wraps that Genz-Sobol backend
in a deterministic selection policy: supported normal problems use several
fixed sequence offsets and unsupported cases use deterministic FORTRAN rather
than falling back to randomized QMC.

The remaining target is not just "use Sobol" or "warp the tails"; it is to add
the surrounding algorithmic pieces that make the randomized Genz-Bretz path so
effective:

- keep the sequential Genz transformation and the same probability interface;
- reorder variables before integration to reduce effective dimension;
- use a small pilot pass to identify conditional tail regions or high-curvature
  coordinates before selecting any refinement grid;
- replace the present fixed sequence offsets with stronger deterministic
  digital shifts when a stability diagnostic is needed;
- use fixed deterministic shifts only as a reproducible stability check, not as
  randomization;
- keep the FORTRAN deterministic path as a reference and fallback.

The fully developed version would aim at the speed of `GenzBretz()` by using
fast point generation, embedded refinement, effective-dimension reduction, and
selective tail-aware refinement. It would aim at the accuracy of
`GenzBretz()` by retaining the Genz transformation and using a stronger
low-discrepancy rule than the current Halton prototype. The important caveat is
that the error attribute would still be deterministic: it could
measure convergence and stability across fixed blocks, but it would not be a
Monte Carlo confidence statement.

For citation purposes, the Sobol-based rows should be tied to Sobol (1967) for
the sequence, Caflisch (1998) or Glasserman (2004, Chapter 6) for QMC
background, and the Genz/Genz-Bretz literature already cited by `mvtnorm` for
the transformed multivariate normal and Student t probability algorithms. The
corresponding BibTeX entries are in `inst/REFERENCES.bib`.

## Accuracy caveats

There are two different notions of uncertainty in this comparison. The
`GenzBretz()` error attribute is tied to randomized integration. The
`DeterministicQMC()` error attribute is a deterministic stability diagnostic.
It can be used to request reproducible convergence checks through `abseps` and
`releps`, but it is not a stochastic confidence interval and it is not a
rigorous deterministic error bound. For applications where formal stochastic
error assessment is important, `GenzBretz()` remains the conservative default.
